---
title: "Chapter 16 Notes"
subtitle: "ggplot2"
author: "Megan Frary"
date: today
toc: true
toc-depth: 2
number-sections: true
format:
  html:
    code-fold: false
---

# Loading Packages

```{r}
# run once
#install.packages("ggplot2")
#| message: false
#| warning: false
library(dslabs)  # Data science labs package
library(dplyr)   # Data manipulation package
library(ggplot2) # Data visualization
```

# What is ggplot2?

```{r}
# ?ggplot()
```

# The Blank Canvas

Load data
```{r}
data("divorce_margarine")  # Load the divorce_margarine data
View(divorce_margarine)    # View the data frame
```

Start every plot by piping a data frame into `ggplot()`

```{r}
divorce_margarine %>% ggplot()
```

# Mapping Aesthetics

Aesthetic mapping is the process of connecting variables in your data to visual properties of the plot, such as position, color, size, and shape. In ggplot2, this is done using the `aes()` function.

```{r}
# ⚡Code Example: Mapping aesthetics
divorce_margarine %>%
  ggplot(aes(x = divorce_rate_maine, y = margarine_consumption_per_capita))
```
🗣 Comment: The plot is still blank because we haven’t added any geometric objects (geoms) yet. The axes are now labeled by the mapped variables, even if there are no points yet. aes() connects data columns to visual positions. The aes() step comes before choosing a geom because we are first defining the mapping (what goes on the x- and y-axes, colors, etc.) and then deciding how to draw those mapped variables (points, lines, bars, etc.).

🪞 Reflection: The aesthetics describe the underlying structure of the plot, and the geoms are just different ways of rendering that structure. By setting aes() inside ggplot() we create a shared mapping that all later geoms can reuse, instead of repeating ourselves for each layer.

Geometric objects (geoms) are the actual shapes drawn on the plot: points, lines, bars, boxplots, etc. Each geom answers the question: Given these aesthetics, how should the data be drawn?

# Scatterplots

🎯 The GOAL: Add points to the plot using geom_point().

```{r}
# ⚡Code Example: Scatterplot
divorce_margarine %>%
  ggplot(aes(divorce_rate_maine, margarine_consumption_per_capita)) +
  geom_point()
```
🎯 The GOAL: Map additional aesthetics inside geom_point(): color by year, size by divorce rate, and transparency (alpha) by margarine consumption.

```{r}
divorce_margarine %>%
  ggplot(aes(x = divorce_rate_maine,   # Global aes mapping: x-axis
             y = margarine_consumption_per_capita)) + # Global aes mapping: y-axis
  geom_point(aes(color = year, size = 20))        # Local aes mapping: color points by year
```
```{r}
# ⚡Code Example: Scatterplot with trend line
divorce_margarine %>%
  ggplot(aes(x = divorce_rate_maine, 
             y = margarine_consumption_per_capita)) +
  geom_point(aes(color = year)) +          # Scatter points
  geom_smooth(method = "lm", se = FALSE)   # Add linear trend line
```


# Histograms

🎯 The GOAL: Simulate a Tallness data frame with heights for 1,000 people in the USA and Canada each. In the column Country, make two groups labeled “USA” and “CAN”. USA heights are centered around 67 inches with a standard deviation of 2.5. CAN heights are centered around 66 inches with a standard deviation of 1.5. Tallness follows a normal distribution.

```{r}
# ⚡Code 
set.seed(2025)
Tallness <- data.frame(
  Country = c(rep("USA", 1000), rep("CAN", 1000)),
  Height  = c(rnorm(1000, 67, 2.5), rnorm(1000, 66, 1.5))
)
```

```{r}
nrow(Tallness)
Tallness %>% 
  group_by(Country)%>%
  summarize(n = n(),
            mean_height = mean(Height),
            sd_height = sd(Height),
            .groups = "drop")
```

```{r}
# ⚡Code Example: Histogram
Tallness %>%
  ggplot(aes(Height)) +    # only x for histograms
  geom_histogram()
```
🎯 The GOAL: Customize the histogram by setting fill, color, and binwidth.

```{r}
# ⚡Code Example: Customized Histogram
Tallness %>%
  ggplot(aes(Height)) +                # only x for histograms
  geom_histogram(fill = "steelblue", 
                 color = "white",      # white border gives gap between bars
                 binwidth = 0.5)       # Set bin width to 0.5
```
## Density plot

```{r}
# ⚡Code Example: Density Plot
Tallness %>%
  ggplot(aes(x = Height)) +
  geom_density(fill     = "purple",
               alpha    = 0.4,
               color    = "black",
               linewidth = 4)
```
## overlaid histograms

```{r}
# ⚡Code Example: Overlaid Histograms
tall_compared_fig <- Tallness %>%
  ggplot(aes(Height, fill = Country)) +
  geom_histogram(position = "identity", 
                 alpha = 0.6)           # Set transparency to see overlap
# ✅ Print the figure
print(tall_compared_fig)
```

# Boxplots

🎯 The GOAL: Build side-by-side boxplots of height by country using geom_boxplot(). Map Country to the x-axis and Height to the y-axis. Use fill = Country to color the boxes by country and set show.legend = FALSE to hide the legend. We’ll store the boxplot as Tall_Box.

```{r}
# ⚡Code Example: Boxplot
Tall_Box <- Tallness %>%
  ggplot(aes(Country, Height, fill = Country)) +
  geom_boxplot(show.legend = FALSE)
# ✅Print the boxplot
Tall_Box
```

# Labels and Themes

We can enhance our plots by adding axis labels and applying themes for better aesthetics.

xlab() and ylab() functions set the labels for the x and y axes, respectively. Simply type the desired label text as a string inside the function.

Additionally, ggtitle() adds a title to the plot.

🎯 The GOAL: Add axis labels and a light theme to the boxplot of height by country.

```{r}
# ⚡Code Example: Boxplot with labels and theme
Tall_Box <- Tallness %>%
  ggplot(aes(Country, Height, fill = Country)) +
  geom_boxplot(show.legend = FALSE) +
  xlab("North American Country") +
  ylab("Height (inches)") +
  ggplot2::theme_light()
# ✅Print the boxplot
Tall_Box
```

