---
title: "Chapter 12 Notes"
subtitle: "mutate() and case_when()"
author: "Megan Frary"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    embed-resources: true
---

# Quarto Notes

# Load Packages
```{r}
#| message: false
#| warning: false
library(dplyr)
library(dslabs)
```

# Vectorized Conditionals
Logical conditions in R can do more than select values. They can also be used to return a new vector by assigning an output to each element based on one or more rules. Functions such as ifelse() and case_when() are called vectorized conditional functions because they evaluate conditions element-by-element across a vector and return a new vector of the same length.

## review

### subsetting with conditions

```{r}
x <- 1:6 # create x
x>3 # logical vector
x[x>3] # logical subsetting
```

### subassignment
```{r}
x[x>3] <- 99 #will assign 99 to all values where x>3
x
```
That is the key new idea in this section.  
* Vectorized conditional functions such as ifelse() and case_when() still test conditions element-by-element, but they do not drop nonmatching cases.  
* Instead, they evaluate a condition for each element and return a new vector with one result for every position.

## ifelse()
goal: Label the vector of numbers greater than 3 as “Large” and others as “Small”.

```{r}
x <- c(3,2,3,6,5)
length(x)
```
`x` is a vector with `{r} length(x)` values.

```{r}
ifelse(test = x > 3, yes = "Large", no = "Small")
length(ifelse(x>3,"Large", "Small"))
```
**Important note:** `ifelse()` returns a NEW vector with a value for every position based on the condition. 

## case_when()
`dplyr` describes `case_when()` as a general vectorized if-else that evaluates rules in order and uses the first match for each element. This makes it especially useful when you have several categories to assign.

```{r}
case_when(c(3, 2, 3, 6, 5) > 3 ~ "Large", .default = "Small")
```
It's interesting how `.default` is used here. Also, now I can see the utility of `case_when()` so that you don't have to next a bunch of if statements!  

## More than Two Categories
```{r}
scores <- c(55, 65, 75, 85, 95)
```

### Nested ifelse()
```{r}
ifelse(scores < 60, "Low",
       ifelse(scores < 80, "Medium", "High"))
```

### Multiple Rule case_when()
```{r}
case_when(scores < 60 ~ "Low",
          scores < 80 ~ "Medium",
          TRUE        ~ "High")
```

# mutate()
`mutate()` lets us store the result of the `case_when()` analysis in a new column! 

`mutate()` creates new columns that are functions of existing columns. It can also modify an existing column if you reuse the same column name.

📜 The SYNTAX: mutate(data, new_col = expression)

Or with piping: name_data <- data %>% mutate(new_col = expression)

* data is the data frame you are working with.  
* new_col is the name of the new column you want to create.  
* expression is the vector or calculation you want to store in that column.

## Column Copy
```{r}
names(mtcars)
```
```{r}
mtcars2 <- mtcars %>% 
  mutate(new_column_copy = am)
names(mtcars2)
```

### mutate() with Arithmetic
```{r}
measurements <- data.frame(height_in = c(60, 65, 70))
measurements
```

The `measurements` data frame has one column: `{r} colnames(measurements)`.

```{r}
measurements %>%
  mutate(height_cm = height_in * 2.54)
```

### case_when() inside mutate()
```{r}
student_scores <- data.frame(scores = c(55, 68, 72, 83, 91))
student_scores <- student_scores %>%
  mutate(
    score_level = case_when(
      scores < 60 ~ "Low",
      scores < 80 ~ "Medium",
      TRUE        ~ "High"
    )
  )
head(student_scores)
```
`student_scores` has the columns `{r} paste(names(student_scores), collapse = " and ")`.

## Overwrite Column Values
```{r}
unique(mtcars$am) 
```
```{r}
cars_overwrite <- mtcars %>%
  mutate(
    am = case_when(
      am == 0 ~ "Automatic", 
      TRUE ~ "Manual")
    )
unique(cars_overwrite$am)
```
The `am` column in `cars_overwrite` has the unique values: `{r} unique(cars_overwrite$am)`.
