---
title: "Chapter 12 Practice"
subtitle: "mutate() and case_when()"
author: "Megan Frary"
date: today
format:  html
---

## Task 0: load packages
```{r}
#| message: false
#| warning: false

library(dslabs)
library(dplyr)
```


## Task 1: concept check

**vectorized conditional functions** like `ifelse()` and `case_when()` return a new vector with one result for every position. 

## Task 2: simple `ifelse()`

**Goal:** Let's create an object called `height_label` using `ifelse()` to label each `height` in the dataset `heights` as "Tall" if `height` is greater than 68 and "Not Tall" otherwise.

```{r}
#?dslabs::heights
heights <- dslabs::heights
names(heights)
```
```{r}
megan_height_label <- ifelse(heights$height>68, "Tall", "Not Tall") # my approach

height_label <- ifelse(
  test = heights$height > 68,
  yes = "Tall",
  no = "Not Tall"
)
```

```{r}
table(height_label)
length(height_label)
nrow(heights)
```
`Table()` seems like a useful function that I don't know that we've encountered yet. It's handy to show the counts like that. 

**Comment:** There are `{r} sum(height_label=="Tall")` tall people and `{r} sum(height_label == "Not Tall")` people. **Note:** I made those numbers show up by using inline code. Claude helped me figure out how to write it! 

**Reflection:** `ifelse()` returned a label for each row. I know this because `length(height_label)` and `nrow(heights)` return the same number which menas they have the same length. 

## Task 3: recode with mutate() and case_when()

**Goal:** We want to label cars based on their mileage as low MPG (mpg < 15), high MPG (mpg > 25), or medium MPG (otherwise).

```{r}
mpg_recode <- mtcars %>%
  mutate(mpg_class=case_when(
    mpg > 25 ~ "High MPG",
    mpg < 15 ~ "Low MPG",
    TRUE ~ "Medium MPG"
  ))
```

```{r}
table(mpg_recode$mpg_class)
head(mpg_recode)
sum(mpg_recode$mpg_class == "Low MPG")
```

**Comment:** The output of table is as expected, with three different labels. There are `{r} sum(mpg_recode$mpg_class == "Low MPG")` cars with low MPG, `{r} sum(mpg_recode$mpg_class == "High MPG")` cars with high MPG, and `{r} sum(mpg_recode$mpg_class == "Medium MPG")` cars with medium MPG.

**Reflection:** By using `mutate()`, we added a new column to the existing dataframe which captures the mileage class of each car. We used `case_when()` to apply labels based on the `mpg` value in the dataset. 

I can see how mutate might be very helpful in the ADATSS reports or just in data analysis more generally! 


## Task 4: recode - name focus

**Goal:** We want to create a new object called `stars_recode` which classifies stars according to their temperature (in a new column called `temp_group`).

**Fun fact:** I spent about 4 years volunteering / working at the Adler planetarium in Chicago and have done lots of activities with kids about sorting stars according to their temperatures. 

```{r}
# ?dslabs::stars
stars <- dslabs::stars

# View(stars)
colnames(stars)
summary(stars$temp)
```
```{r}
stars_recode <- stars %>%
  mutate(
    temp_group = case_when(
    temp < 4000 ~ "Cool",
    temp <= 7500 ~ "Warm",
    TRUE ~ "Hot"
  ))
table(stars_recode$temp_group)
head(stars_recode)
```

**Comment:** The column `temp` shows the temperature values but the column `temp_group` sorts them as cool, warm, or hot (relatively speaking).   The new column goes on the left side of the = sign. 

**Reflection:** The order of the `case_when()` rules matter here because if we'd put the warm rule first then even the cool stars would be classified as warm. Although maybe not! I just wrote the warm rule as <= 7500, but we could have written the warm rule with an and statement (>4000 & <=7500) and then we could have put it first and it would have worked just fine. 


## Task 5: unit conversion 

**Goal:** We want to create `international_friendly_heights` from the `height` column of `heights` by overwriting the value that is in inches with one that is in centimeters. 

```{r}
heights <- dslabs::heights
head(heights)
summary(heights$height)
```


```{r}
internationally_friendly_heights <- heights %>%
  mutate(height = height* 2.54)
head(internationally_friendly_heights)
summary(internationally_friendly_heights$height)
```

**Comment:** This code overwrote the column because we used an existing column name. If we'd given it a new name, then it would have added on an additional column. I know because I used `head()` to look at the table structure. 

**Reflection:** The new height values look reasonable for cm. In the new dataframe, the height values are now given in cm instead of inches.

I can see how this replacement approach could be useful in the ADATSS data (if you were a confident coder!). For example, maybe we want to replace letter grades with grade groups. We wouldn't use this 1:1 replacement, but would use case_when() and could change a column that isn't useful in its current format into one that is. 
