---
title: "Chapter 11 Notes"
subtitle: "dplyr Subsetting"
author: "Megan Frary"
date: today
format:
  html:
    code-fold: false
execute:
  warning: true
  message: true
---

# Quarto
```{r}
#| eval=FALSE
# just some test code that will not run
x <- 1.5
mean(x)
```

# Packages

The `dplyr` package is part of the **tidyverse**, a collection of R packages designed for data science. The `dplyr` package gives you data-wrangling functions called **verbs**. These verbs help you work with data frames in readable steps. In this chapter, we will focus on `select()` for choosing columns and `filter()` for keeping rows.

```{r}
#| eval=FALSE

# Install packages
install.packages("dslabs")
install.packages("dplyr")
```

```{r}
#| message=FALSE
library(dslabs)
library(dplyr)
```

# The Pipe
```{r}
mtcars %>% head()
```

```{r}
mtcars %>%
  head()
```

# Selecting Columns

## positive selection
```{r}
select(mtcars, mpg, cyl, hp)    # keep three columns
```

## negative selection
```{r}
select(mtcars, -gear)   # drop one column
```

## piping with select

```{r}
mtcars %>%
  select(mpg,cyl,hp) # easier to read? I say "not really"
```

## subset with select
```{r}
I_Demo <- gapminder %>%
  select(country,year,infant_mortality, continent)
head(I_Demo)
```
```{r}
# test this myself
names(gapminder)
```
```{r}
Your_Choice <- gapminder %>%
  select(fertility, population, gdp)
head(Your_Choice)
```

# Filtering Rows
`filter()` keeps only rows for which the condition is `TRUE`. That means NA values aren't kept! 

## single condition
```{r}
filter(mtcars,mpg >= 25) #high mileage cars
```
## piping with filter
```{r}
nrow(murders)
```
```{r}
unique(murders$state)
```
```{r}
No_Washington <- murders %>% 
  filter(state != "District of Columbia")
nrow(No_Washington)
```
```{r}
unique(No_Washington$state)
```
## multiple conditions with AND
```{r}
filter(mtcars, mpg >= 25, cyl == 4)
```
```{r}
filter(mtcars, mpg >= 25 & cyl == 4)
```

### with the pipe
```{r}
efficient <- mtcars %>%
  filter(mpg >= 25, cyl == 4) # two conditions
nrow(efficient) # how many cars are efficient? 
```
## multiple conditions

### or | and %in%
```{r}
filter(mtcars, cyl == 4 | cyl == 6)
```
```{r}
filter(mtcars, cyl %in% c(4, 6))
```
```{r}
filter(mtcars, cyl == 4 | mpg >= 30)  # different variables
```
### with the pipe
```{r}
efficient_ish <- mtcars %>%
  filter(mpg >= 25 | cyl == 4)    # two conditions
nrow(efficient_ish) # how many cars are efficient_ish?
```

# filter() and select()
🎯 Using us_contagious_diseases from dslabs, keep only rows for Alaska from 1980 and later. Then keep only the columns year, disease, and count. Store the result as alaska_disease.
```{r}
alaska_disease <- us_contagious_diseases %>%
  filter(state == "Alaska",year >=1980) %>%
  select(year, disease, count)
head(alaska_disease)
```

# Application
❓ Which automatic cars (am == 0) in mtcars have horsepower greater than 150?

```{r}
powerful_auto <- mtcars %>%
  filter(am ==0, hp>150) %>%
  select(hp,mpg)
head(powerful_auto)
nrow(powerful_auto)
```

