---
title: "Chapter 11 Practice"
author: "Megan Frary"
date: today
format:
  html:
    code-fold: false
execute:
  warning: true
  message: true
---

## Task 0: Load packages

**Goal:** Load the packages that we'll need: dplyr and dslabs

```{r}
#| message: false
#| warning: false

library(dslabs)
library(dplyr)
```

**Comment:** Even though these packages have already been installed, we still need to use library() to basically load them in to this notebook. 

**Reflection:** It might be useful to show messages and warning while you're learning so that you see what's happening in RStudio. However, it's useful to hide them in a final report so that it looks nicer. 

## Task 1: Column selection

**Goal:** We want a new data frame called C_overload that keeps only the year and carbon_emissions column from the temp_carbon data frame. We'll use select without piping. 

```{r}
C_overload <- select(temp_carbon, year, carbon_emissions)
head(C_overload)
```

**Comment:** We used select to keep only certain columns. Without piping, the first argument needed to be the dataset, then I named the columns I wanted to keep. head() helps me make sure that the new dataset looks as expected. 

**Reflection:** I think that select is easier because you don't have to use "" around the column name! I was always forgetting that anyway. We could use head() or colnames() to verify the output. 

## Task 2: single-column dataframe 

**Goal:** Create a data frame named star_names that uses the stars dataset (from dslabs), keeping only the star column. Use piping and select(). 

```{r}
star_names <- stars %>% select(star)
head(star_names)
```

**Comment:** Given the output of head(), it shows that this is a df[6x1] so that tells me this is a dataframe and not a vector. 

**Reflection:** We need to know if something is a vector or one-column data frame because they will work differently. For example, I think that we can't take the mean of a one-column dataframe the same way we could if it were a vector. 

## Task 3: row filtering

**Goal:** Create a new object called No_Florida that removes the florida rows from the murder dataset. We'll use !=

```{r}
unique(murders$state)
```
```{r}
No_Florida <- murders %>% filter(state != "Florida")
unique(No_Florida$state)
```

**Comment:** It's helpful to pre-check the states so that you know what options you have. When we verify, we expect to see the same list but without Florida in it. 

**Reflection:** I first wrote the code without looking and didn't use piping. This is a short code so either way would work. 

## Task 4: multiple conditions 

**Goal:** We're making a new object called Fast_n_Furious that uses mtcars and selects for those that are manual, have 8 cylinders, and have 1/4 mile time less than 15

```{r}
# Fast_n_Furious <- mtcars %>%
#  select(am == 1, cyl == 8, qsec < 15)
# head(Fast_n_Furious)
# nrow(Fast_n_Furious)
```
Here's the error I got (I wrote the code without looking at the example code)
Error in `select()`:
ℹ In argument: `am == 1`.
Caused by error:
! object 'am' not found
Run `rlang::last_trace()` to see where the error occurred.

```{r}
unique(mtcars$am)
Fast_n_Furious <- mtcars %>% 
  filter(am == 1, cyl ==8, qsec < 15)
head(Fast_n_Furious)
nrow(Fast_n_Furious)
```
**My mistake** was to use select (which keep certain columns) instead of filter (which keeps certain rows). I reread the instructions and saw that that was specified there. 

**Comment:** The way we used filter implied AND logic. There are only two cars which meet all three conditions: they are manual, they have 8 cylinders, and they need < 15 s to go a 1/4 mile. 

**Reflection:** If the result wasn't as expected (although that implies that I knew what to expect from this dataset), I guess I would try each condition at a time and look at how the down-selecting is working. 

## Task 5: OR logic + column slice

**Goal:** Create a new object called filtered_cars which uses mtcars and keeps only those whose displacements is < 160 or >250 and then shows only the firs three columns of those rows. 
```{r}
range(mtcars$disp)
```
**Comment:** Using range shows us what the full range is so that we can make decisions about the cutoff values. 

```{r}
filtered_cars <- mtcars %>%
  filter(disp>250 | disp<160) %>%
  select(1:3)
all(filtered_cars$disp > 250 | filtered_cars$disp < 160) # check disp condition
```

**Comment:** Ha! I didn't know what all() did but mine gave back FALSE to start with. Then I read about all in the comments and see that it's just a logic test to see if the values meet the logic conditions. Because the instructions said it should be true, that prompted me to revisit my original code. I saw that I had put >160 instead of <160 so I fixed that. all() saved the day (and reminded me of the importance of using some kind of check when we've created a new dataset).

**Reflection:** We need OR here because a car can't actually meet both of these conditions at the same time. We want those that meet either condition and that's OR logic. 

## Task 6

**Comparing selection methods** We've learned a few different ways to subset, first using [] and then with filter() and select(). To me they are equally easy to read because I know what they are both meant to do. However, maybe using the names filter() and select() remind us of what we're doing in ways that are helpful. I also appreciate that filter() and select() don't need "" around the column names, an error I'm prone to making. They seem equally easy to verify; for example, I think we could use all(), as we did in task 5, to verify regardless of how we had created the new filtered_cars object. In fact, let's try! 

```{r}
#megan_filtered_cars <- mtcars["disp"<160 | "disp">250,1:3] first attempt: WRONG! 
megan_filtered_cars <- mtcars[mtcars$disp<160 | mtcars$disp >250,1:3]
head(megan_filtered_cars)
identical(filtered_cars,megan_filtered_cars)
```
Well, maybe I'm wrong about using "" for column names, at least in that way. But you do still have to refer to the columns differently (which I figured out). So there are two ways to do it, but I think filter and select are preferred. 

**Reflection:** I don't know that I know yet when to choose base R brackets and when to use dplyr verbs. I think I need more practice looking at the two cases to see which would be better. I asked Claude to suggest ideas about when each was preferred. A few of the ideas that it provided about using dplyr verbs made sense to me, especially in light of the ADATSS code. These included: (1) when using complex, multi-step analysis (this makes sense that you'd sort of chop up the dataset in parts and the dplyr verbs woudl help), (2) data cleaning and transformation (this example included mutate() which we haven't learned yet but which I think I saw in the ADATSS code), (3)
 when readability matters (I know you wrote the ADATSS code with this in mind which is probably why you opted for dplyr verbs there), and (4) for modern R workflows that use the tidyverse ecosystem (although we haven't done that here, I know you use ggplot2 in the ADATSS report so that might be another reason). 

