---
title: "Chapter 13 Notes"
subtitle: "group_by() and summarize()"
author: "Megan Frary"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    embed-resources: true
---

# Loading Packages
```{r}
#| message: false
#| warning: false
library(dslabs)  # Data science labs package
library(dplyr)   # Data manipulation package
```

# count()
```{r}
mtcars %>%
  count(cyl)
```

### count() vs length()
length() counts how many values are in a vector.  
count() counts rows in a data frame for each category or combination of categories.

### count() with sorting
Count cars by cylinder and sort from largest group to smallest group. `sort = TRUE` arranges the result so the largest count appears first. The counts are the same as before; only the row order changed.
```{r}
mtcars %>%
  count(cyl, sort = TRUE)
```

## count multiple variables
Count how many cars fall into each combination of cylinder count and transmission type.
```{r}
mtcars %>%
  count(cyl, am)
```
If our question is only “How many rows are in each category?”, `count()` is enough. But what if our question asks about averages, maximums, spread, or other statistics? We’ll need `summarize()`.

# summarize()
```{r}
mtcars %>% 
  summarize(mean_mpg = mean(mpg))
```

```{r}
mtcars %>% 
  summarize(mean_mpg = mean(mpg),
            max_mpg = max(mpg),
            count = n())
```
`summary()` is a base R function that gives a quick overall description of an object, such as basic statistics for each variable in a data frame. `summarize()` from `dplyr` is used to create your own summary results, usually after `group_by()`, such as the mean or count for each group.

## count() vs summarize()
```{r}
mtcars %>%
  group_by(cyl) %>%
  summarize(n = n())
```

# group_by()
`group_by()` changes the behavior of summary functions. 🗣 `group_by(cyl)` splits the cars into groups based on cylinder count, and `summarize()` calculates `mean(mpg)` for each group. `mean_mpg` is the name of the new summary column.

```{r}
cyl_means <- mtcars %>%
  group_by(cyl) %>%
  summarize(mean_mpg = mean(mpg))
cyl_means
```

## explicit vs default naming
```{r}
mtcars %>%
  group_by(cyl) %>%
  summarize(mean(mpg))      

```

# Multiple Statistics
```{r}
cyl_stats <- mtcars %>%
  group_by(cyl) %>%
  summarize(mean_mpg = mean(mpg),
            sd_mpg   = sd(mpg),
            n        = n())
cyl_stats
```

# Multiple Variables
`group_by()` with one column creates one group for each unique value in that column. `group_by()` with two columns creates one group for each unique combination of values.

```{r}
sort(unique(mtcars$cyl))
sort(unique(mtcars$am))
```
```{r}
mtcars %>%
  group_by(cyl, am) %>%
  summarize(avg_mpg = mean(mpg),
            count = n())
```
## ungroup

```{r}
mtcars %>%
  group_by(cyl, gear) %>%
  summarize(mean_mpg = mean(mpg))   # no .groups set
```

```{r}
cyl_gear_stats <- mtcars %>%
  group_by(cyl, gear) %>%
  summarize(mean_mpg = mean(mpg))   # no .groups set
```
```{r}
group_vars(cyl_gear_stats)  # what groups remain?
```
```{r}
is_grouped_df(cyl_gear_stats) # expect TRUE
```
```{r}
cyl_gear_stats_drop <- mtcars %>%
  group_by(cyl, gear) %>%
  summarize(mean_mpg = mean(mpg), .groups = "drop")
```

```{r}
group_vars(cyl_gear_stats_drop)  # expect character(0)
is_grouped_df(cyl_gear_stats_drop) # expect FALSE
```
```{r}
cyl_gear_stats_ungroup <- mtcars %>%
  group_by(cyl, gear) %>%
  summarize(mean_mpg = mean(mpg)) %>%
  ungroup()
```

